Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-05 main.tex
Edits to last parts of the intro for better wording.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with steer motor and data
    acquisition and control electronics mounted in the rear rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:

\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \mathbf{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0
\quad k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts
for \(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to
\(v\), the speed range where the bicycle is stable can be maximized within any
physical actuator magnitude and bandwidth limits.  \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller on this model.

\todo[inline]{I'm not consistent with the arrow being a vector for $B_\delta$}

In this paper, we test whether a gain scheduled steer motor controlled bicycle
that is stable at a low speed range when riderless assists in balancing by
investigating whether it is beneficial in preventing the rider from falling.
To do this, we apply varying magnitude mechanical perturbations to the
handlebars while the rider is balancing the bicycle on a treadmill and then
assess the rider's probability of falling with the balance assist steer motor
system on and off.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

@@

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:

\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \vec{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{min} < v < v_{capsize}\). With gain scheduling with respect to \(v\), the
speed range where the bicycle is stable can be maximized within any physical
actuator magnitude and bandwidth limits. \citet{Schwab2008} elaborates on some
of the possibilities in scheduling the gains for such a controller and shows
that a linear scheduling with respect to speed can give satisfactory stability.

In this paper, we test whether a gain scheduled steer motor controlled bicycle
that is stable in a large low speed range (when riderless) assists the rider in
balancing by investigating whether it is beneficial in preventing falling.  To
do this, we apply varying magnitude mechanical perturbations to the handlebars
while the rider is balancing the bicycle on a treadmill and then assess the
rider's probability of falling with the balance assist steer motor system on
and off.
